### 2019-07-14 mathematical-background/information-theory.tex
mutual information section

>>>>

\subsection{Mutual Information} \label{subsec:mutualinformation}
Another way of characterising the interrelatedness of two variables is through the concept of \textit{mutual information}, as defined in \cite{Cover2006}, that is closely linked to entropy (Eq. \ref{eq:entropy}).
\begin{definition}
	The mutual information of two random variables $X$ and $Y$ is given by:
	\begin{equation} \label{eq:mutual-information}
		I(X,Y) = \sum_{x \in \mathcal{X}} \sum_{y \in \mathcal{Y}} p_{XY}(x,y) \log \left( \frac{p_{XY}(x, y)}{p_{X}(x) p_{Y}(y)} \right)
	\end{equation}
\end{definition}
NB: In Information Theory, the convention is that $0 \log(0) = 0$.

$I_{XY}$, intuitively, measures the amount of information that $X$ and $Y$ share that can also be seen as the degree to which one variable is informative of the other.
If $X$ and $Y$ are independent then they share no mutual information and knowing one of the two gives no information about the other.
This can be immediately understood by rewriting Eq. \ref{eq:mutual-information} as:
\begin{equation}
	I(X,Y) = H(X) - H(X|Y) = H(Y) - H(Y|X)
\end{equation}
The mutual information I(X,Y) is the reduction in uncertainty of one of the variables given the knowledge of the other.
if $X$ and $Y$ are perfectly correlated ($\rho_{XY}= \pm 1$) then they both convey the same amount of information and $I_{XY}$ is equal to the entropy $\Eta(X) = \Eta(Y)$.

### 2019-07-14 methodology/instruments.tex
Algorithmicx package
mutual information section

\section{Methods} 
\subsection{Libraries}\label{subsec:libraries}
\subsubsection{Pomegranate}
\textit{pomegranate} (\cite{pomegranate}) is an open-source probabilistic models package for python.
Its core philosophy is that every probabilistic model, from Hidden Markov to Bayesian Network, can be seen as a probability distribution and, as such, can be flexibly composed into hierarchical mixture models (\cite{Schreiber2017}).
The package implements:
\begin{itemize}
	\item Probability Distributions
	\item General Mixture Models
	\item Hidden Markov Models
	\item Bayes Classifiers and Na{\"i}ve Bayes
	\item Markov Chains
	\item \textbf{Bayesian Networks}
	\item Factor Graphs
\end{itemize}

@@

\textit{Structure learning} from data is achieved using the \texttt{from\_samples} method of the \texttt{BayesianNetwork} class, with the default algorithm being the novel one described by \cite{schreiber_noble_2017}.
The \textit{probability} of a sample is calculated using the \texttt{probability} function of an object of \texttt{BayesianNetwork} type; the \texttt{predict\_proba} function is used to return the probability of each variable in the model given some evidence.
\textit{Predictions} (described in detail in Sec. \ref{subsec:bnupdating}) are run by passing to the \texttt{predict} function of an object a matrix with \texttt{None} as placeholders for missing values .
\textit{Fitting} is done thought the \texttt{fit} function that uses MLE estimates to update each node's distribution in the model based on the input data.

@@

@@

\subsubsection{Networkx}
\textit{NetworkX} (\cite{networkx}) is another widely-used package that is specialised in the creation and manipulation of graph-structured data.
The main use for this package was in building the \texttt{MPEGraph} ( \todo[inline]{inserire referenza una volta scritto} ) data structure that I used to build a graph representing the outcome of pseudo-MPE (see \ref{subsubsec:pseudo-mpe} ) applied to multiple initial evidences (see \ref{subsubsec: pseudo-mpe-random} ) .

\subsubsection{Anytree}
\textit{anytree} (\cite{anytree}) is a python package providing a lightweight implementation of a tree data structure and traversal methods.
This package was used as a lightweight substitute for NetworkX when building a chain of reasoning using the pseudo-MPE algorithm (see \ref{subsubsec:pseudo-mpe} ).
As the \texttt{MPETree} object is guaranteed to be a single chain, using a simpler resource than NetworkX was advantageous.

\subsection{Algorithms}
\todo{quanto dettagli implementativi devo mettere?  perchè ho un sacco di funzioni di visualizzazione etc}

@@

\subsubsection{d-separation}
A na{\"i}ve implementation according to the definition (presented in Subsec. \ref{subsec:bayesiannetworks}) to check for d-separation between node $X$ and $Y$ would have a complexity in the order of the number of trails between $X$ and $Y$; this leads to an exponential in the size of the graph running time.
Luckily, \cite{koller2007dseparation} present what is a linear time algorithm to solve the problem.

@@

\begin{algorithm}[htp!]
	\caption{d-separation algorithm}
	\label{alg:d-separation}
	\begin{algorithmic}
		\STATE {\bf input:} source $X$, evidence $E$, nodes $V$
		\STATE $separated\_list = \emptyset$
		\FOR{target $Y \in V$} 
		\STATE append $d-separated(X, Y, E)$ to $separated\_list$ \COMMENT{will return true or false}
		\ENDFOR
		\STATE {\bf output:} $separated\_list$
	\end{algorithmic}
\end{algorithm}

>>>>

\section{Methods}

\subsection{Libraries}\label{subsec:libraries}
\subsubsection{Pomegranate}
\textit{pomegranate} (\cite{pomegranate}) is an open-source probabilistic models package for python.
Its core philosophy is that every probabilistic model, from Hidden Markov to Bayesian Network, can be seen as a probability distribution and, as such, can be flexibly composed into hierarchical mixture models (\cite{Schreiber2017}).
The package implements:
\begin{itemize}
	\item Probability Distributions
	\item General Mixture Models
	\item Hidden Markov Models
	\item Bayes Classifiers and Na{\"i}ve Bayes
	\item Markov Chains
	\item \textbf{Bayesian Networks}
	\item Factor Graphs
\end{itemize}

@@

\textit{Structure learning} from data is achieved using the \texttt{from\_samples} method of the \texttt{BayesianNetwork} class, with the default algorithm being the novel one described by \cite{schreiber_noble_2017}.
The \textit{probability} of a sample is calculated using the \texttt{probability} function of an object of \texttt{BayesianNetwork} type; the \texttt{predict\_proba} function is used to return the probability of each variable in the model given some evidence.
\textit{Predictions} (described in detail in Subsec. \ref{subsec:bnupdating}) are run by passing to the \texttt{predict} function of an object a matrix with \texttt{None} as placeholders for missing values .
\textit{Fitting} is done thought the \texttt{fit} function that uses MLE estimates to update each node's distribution in the model based on the input data.

@@

\subsubsection{Pgmpy}
\textit{Pgmpy} (\cite{pgmpy}) is, like pomegranate, another recent probabilistic graphical model package for Python.
Unlike pomegranate, it natively implements various exact and approximate inference algorithms, like Variable Elimination, Belief Propagation and Max-Product Linear Programming.

The reason that I elected to use two different probabilistic graphical model libraries is because there is still no Python package that offers all the needed functionality.
Pomegranate implements a novel structure learning algorithm, that I wanted to use, but is severely lacking in functionality in many other areas.
Pgmpy, on the other hand, has a very good API as regards inference.

@@

\subsubsection{Networkx}
\textit{NetworkX} (\cite{networkx}) is another widely-used package that is specialised in the creation and manipulation of graph-structured data.
The main use for this package was in building the \texttt{MPEGraph} data structure that I used to build a polytree representing the dialogue with the expert.

\subsection{Algorithms} \todo{thresholded dialogue aggiungere}
\todo{quanti dettagli implementativi devo mettere?  perchè ho un sacco di funzioni di visualizzazione etc}

@@

\subsubsection{d-separation}
A na{\"i}ve implementation according to the definition (presented in Subsec. \ref{subsec:d-separation}) to check for d-separation between node $X$ and $Y$ would have a complexity in the order of the number of trails between $X$ and $Y$; this leads to an exponential in the size of the graph running time.
Luckily, \cite{koller2007dseparation} present what is a linear time algorithm to solve the problem.

@@

\begin{algorithm}[htp!]
	\caption{d-separation algorithm}
	\label{alg:d-separation}
	\begin{algorithmic}[1]
		\State $separated\_list = \emptyset$
		\For{target $Y \in V$} 
			\State append $d-separated(X, Y, E)$ to $separated\_list$ \Comment{will return true or false}
		\EndFor
	\end{algorithmic}
\end{algorithm}
